---
title: "Project A – Blood Pressure Analysis and Statistical Interpretation"
author: "Mohamed Hussein"
date: " 2025-08-20"
output: pdf_document
---

## Objective

This R project analyzes a blood pressure dataset to study how **dose** and **gender** affect **blood pressure reduction**.  
The workflow includes descriptive statistics, visualizations, normality and variance testing, confidence interval estimation, hypothesis testing, and linear regression modeling.  

**Workflow Overview:**
1. Load required R packages.  
2. Import and inspect the dataset.  
3. Perform descriptive and visual statistical analyses.  
4. Assess data normality and homogeneity of variance.  
5. Conduct hypothesis tests and ANOVA.  
6. Build and interpret a linear regression model.

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width=6, fig.height=4)
```
## 1.0 Load Packages

```{r load-packages}
library(googledrive)   # For downloading data directly from Google Drive
library(dplyr)         # For data manipulation (filtering, grouping, summarise)
library(ggplot2)       # For creating plots and visualizations
library(psych)         # For detailed descriptive statistics (describe())
library(car)           # For regression diagnostics and tests (leveneTest, etc.)
library(Hmisc)         # For additional statistical utilities and summaries

```

---

## 2.0 Download and Load Dataset

### 2.1 Access and load dataset

```{r download-data}
drive_deauth()                                                              # Disable Drive auth to access public/shared file
file_id <- "1mGrXc0VuNndsqbR6GKhLLv7jkTQ20Ubn"                              # Google Drive file ID for dataset
drive_download(as_id(file_id), path = "BloodPressure.RData", overwrite = TRUE)  # Download dataset to local file
load("BloodPressure.RData")                                                 # Load RData file into R environment (creates BloodPressure object)

```

---

## 3.0 Explore Data

### 3.1 Inspect structure and summary

```{r explore-data}
str(BloodPressure)    # Show structure: variables, types, and sample values
summary(BloodPressure) # Provide summary statistics (min, 1st Qu., median, mean, etc.)
head(BloodPressure)    # Display first 6 rows for quick inspection
names(BloodPressure)   # List column names for reference

```

---

## 4.0 Descriptive Statistics

### 4.1 Basic summaries and correlations

```{r descriptive}
describe(BloodPressure)                                    # Detailed descriptive stats (mean, sd, skew, kurtosis)
cor(BloodPressure$bp.reduction, BloodPressure$dose)        # Pearson correlation between dose and BP reduction
table(BloodPressure$gender)                                # Frequency table for gender variable
```

---

## 5.0 Graphics and Visualization

### 5.1 Bar chart – gender distribution

                                         
```{r graphics}
ggplot(BloodPressure, aes(x = gender)) +                   # Start ggplot mapping gender to x-axis
  geom_bar(fill = "steelblue") +                           # Draw bar chart with blue fill
  labs(title = "Gender Distribution") +                    # Add plot title
  theme_minimal()                                          # Apply a clean minimal theme

```


### 5.2 Bar chart – mean BP reduction by gender

```{r}
ggplot(BloodPressure, aes(x = gender, y = bp.reduction)) + # Map gender to x and mean BP reduction to y
  stat_summary(fun = mean, geom = "bar", fill = "lightgreen") +  # Plot mean as bars with green fill
  labs(title = "Mean Blood Pressure Reduction by Gender") +      # Title explaining the plot
  theme_minimal()                                                 # Minimal theme for clarity
```


### 5.3 Histogram – dose distribution

```{r}
ggplot(BloodPressure, aes(x = dose)) +                    # Map dose to x-axis
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +  # Histogram with specified bin width and colors
  labs(title = "Histogram of Dose") +                     # Title for the histogram
  theme_minimal()                                         # Minimal theme for consistency

```

### 5.4 Histogram – BP reduction distribution

```{r}
ggplot(BloodPressure, aes(x = bp.reduction)) +            # Map BP reduction to x-axis
  geom_histogram(binwidth = 5, fill = "orange", color = "black") +  # Histogram with larger bin width
  labs(title = "Histogram of BP Reduction") +             # Title explaining the distribution plotted
  theme_minimal()                                         # Minimal theme to match other plots
```


### 5.5 Scatterplot – dose vs BP reduction by gender

```{r}
ggplot(BloodPressure, aes(x = dose, y = bp.reduction, color = gender)) +  # Map dose to x, BP reduction to y, color by gender
  geom_point() +                                                         # Add scatter points
  geom_smooth(method = "lm", se = FALSE) +                               # Add linear regression lines for each gender without CI shading
  labs(title = "Scatterplot Dose vs BP Reduction by Gender") +           # Title for the scatterplot
  theme_minimal()                                                        # Keep visual style consistent
```


### 5.6 Boxplots – distribution overview


```{r}
# Boxplot of BP reduction overall
ggplot(BloodPressure, aes(y = bp.reduction)) +           # Map BP reduction to y-axis for boxplot
  geom_boxplot(fill = "pink") +                          # Draw boxplot with pink fill
  labs(title = "Boxplot of Blood Pressure Reduction") +  # Title to show purpose of plot
  theme_minimal()                                        # Minimal theme for readability

# Boxplot of BP reduction per dose group
ggplot(BloodPressure, aes(x = as.factor(dose), y = bp.reduction)) +  # Treat dose as categorical for group boxplots
  geom_boxplot(fill = "lightblue") +                                   # Fill color for clarity
  labs(title = "BP Reduction per Dose", x = "Dose (mg)") +             # Title and x-label
  theme_minimal()                                                       # Consistent styling
```

---

## 6.0 Outlier Detection

### 6.1 Detect outliers visually

```{r outliers}
boxplot(BloodPressure$bp.reduction, main = "Outlier Detection for BP Reduction")  # Boxplot highlights potential outliers for BP reduction

```

---

## 7.0 Normality and Homoscedasticity

### 7.1 Test for normal distribution

```{r normality}
shapiro.test(BloodPressure$bp.reduction)          # Shapiro-Wilk test for normality of BP reduction
qqPlot(BloodPressure$bp.reduction)                # Q-Q plot to visually assess normality (from car package)

```


### 7.2 Test variance equality (Levene’s test)


```{r homoscedasticity}
leveneTest(bp.reduction ~ as.factor(dose), data = BloodPressure)  # Levene's test for homogeneity of variances across dose groups
plot(bp.reduction ~ dose, data = BloodPressure)                  # Scatter plot to visually inspect variance across doses

```

---

## 8.0 Confidence Intervals per Dose

### 8.1 Calculate 90%, 95%, and 99% CI for mean BP reduction


```{r confidence-intervals}
BloodPressure %>%
  group_by(dose) %>%                                             # Group data by dose level
  summarise(
    mean_bp = mean(bp.reduction),                                # Compute mean BP reduction within dose group

    # 90% Confidence Interval
    ci90_lower = mean_bp - qt(0.95, n()-1) * sd(bp.reduction)/sqrt(n()),  # Lower bound for 90% CI
    ci90_upper = mean_bp + qt(0.95, n()-1) * sd(bp.reduction)/sqrt(n()),  # Upper bound for 90% CI

    # 95% Confidence Interval
    ci95_lower = mean_bp - qt(0.975, n()-1) * sd(bp.reduction)/sqrt(n()), # Lower bound for 95% CI
    ci95_upper = mean_bp + qt(0.975, n()-1) * sd(bp.reduction)/sqrt(n()), # Upper bound for 95% CI

    # 99% Confidence Interval
    ci99_lower = mean_bp - qt(0.995, n()-1) * sd(bp.reduction)/sqrt(n()), # Lower bound for 99% CI
    ci99_upper = mean_bp + qt(0.995, n()-1) * sd(bp.reduction)/sqrt(n())  # Upper bound for 99% CI
  )                                                              # Summarise returns a tibble with mean and CI columns


```

---

## 9.0 Hypothesis Testing

### 9.1 Compare BP reduction between genders at dose = 0


```{r hypothesis-testing}
t.test(bp.reduction ~ gender, data = subset(BloodPressure, dose == 0), var.equal = TRUE)  # Independent t-test for gender difference at dose 0

```

### 9.2 Compare BP reduction between dose 0 and 10 groups


```{r}
t.test(bp.reduction ~ as.factor(dose), data = subset(BloodPressure, dose %in% c(0,10)), var.equal = FALSE, alternative = "greater")  # One-sided Welch t-test: H1 = dose10 > dose0

```


### 9.3 Perform one-way ANOVA across all dose groups


```{r}
anova_model <- aov(bp.reduction ~ as.factor(dose), data = BloodPressure)  # Fit one-way ANOVA model to test mean differences across doses
summary(anova_model)                                                      # Display ANOVA table (F-statistic, p-value)
TukeyHSD(anova_model)                                                     # Post-hoc Tukey HSD for pairwise comparisons between dose levels

```

---

## 10.0 Linear Regression Analysis

### 10.1 Build and interpret linear model


```{r linear-regression}
BloodPressure$gender <- factor(BloodPressure$gender, levels = c("0","1"), labels = c("male","female"))  # Recode gender to descriptive factor labels

lm_model <- lm(bp.reduction ~ dose + gender, data = BloodPressure)   # Fit linear regression predicting BP reduction from dose and gender
summary(lm_model)                                                    # Show regression coefficients, p-values, and model fit
confint(lm_model, level = 0.95)                                      # 95% confidence intervals for coefficients
predict(lm_model, newdata = data.frame(dose = 3, gender = "male"), interval = "confidence")  # Predict mean BP reduction for male at dose=3 with CI


```

---

