
---
title: "Assignment 1 – Fundamental R Programming and Gene Data Analysis"
author: "Mohamed Hussein"
date: " 2025-08-09"
output: pdf_document
---

## Objective

This R notebook demonstrates foundational R programming concepts such as vector creation, data subsetting, basic file handling, and simple data manipulation.  
The goal is to practice essential R operations including indexing, sequencing, descriptive analysis, and vectorized computations using gene-related data.

**Workflow Overview:**
1. Create and manipulate vectors.  
2. Subset and transform elements in numeric sequences.  
3. Explore files in the working directory.  
4. Perform simple bioinformatics-style data operations on gene expression and gene length. 5. Compute derived metrics such as normalized gene expression.  


---

## Exercise 1 – Creating Vectors

### 1. Create a vector containing the values 1,2,3,4,5

```{r}

vec1 <- c(1,2,3,4,5)     # Create a vector with elements 1–5
print(vec1)              # Display the vector

```
### 2. Create a vector containing the values 1 to 100

```{r}
vec2 <- 1:100            # Generate a sequence from 1 to 100
print(vec2)              # Display the sequence


```


### 3. Create a vector containing the values 0,5,10,15,20

```{r}
vec3 <- seq(0, 20, by=5) # Create a sequence with step size 5
print(vec3)              # Display the sequence
```


### 4. Create a vector containing the values 1,1,2,2,3,3


```{r}
vec4 <- c(1,1,2,2,3,3)   # Create a repeated pattern vector
print(vec4)              # Display the vector
```


### 5. Create a vector containing the values 1,1,5,7,9,10



```{r}
vec5 <- c(1,1,5,7,9,10)  # Create a custom numeric vector
print(vec5)              # Display the vector
```
---

## Exercise 2 – Subsetting Vectors

### 1. Create a vector containing the values 1 to 10

```{r}
v <- 1:10                # Create a numeric sequence from 1 to 10
print(v)                 # Display the vector
```


### 2. Create a new vector with all but the first and last value

```{r}
v2 <- v[-c(1,length(v))] # Remove first and last elements
print(v2)                # Display the new vector
```

### 3. Create a new vector with all but the second and fifth value

```{r}
v3 <- v[-c(2,5)]         # Remove elements at positions 2 and 5
print(v3)                # Display the new vector
```


### 4. Create a new vector of square root of the sixth and seventh position

```{r}
v4 <- sqrt(v[c(6,7)])    # Compute square roots of positions 6 and 7
print(v4)                # Display the results

```

### 5. Create a new vector of alternating positions in the vector using another vector

```{r}
indices <- seq(1, length(v), by=2) # Create sequence of odd indices
v5 <- v[indices]                   # Select alternating elements
print(v5)                          # Display the alternating elements

```

---

## Exercise 3 – Working with Files

### 1. Find the number of files in the present working directory

```{r}
num_files <- length(list.files())  # Count files in working directory
print(num_files)                   # Display the count

```

### 2. List the first file in the present working directory

```{r}
first_file <- list.files()[1]      # Get the first file name
print(first_file)                  # Display the first file

```

---

## Exercise 4 – Gene Expression Data

### 1. Create a vector of the gene names Gene_1, Gene_2, Gene_3, Gene_4

```{r}
genes <- c("Gene_1", "Gene_2", "Gene_3", "Gene_4")  # Define gene names
```


### 2. Create a vector of the expression values 1000, 3000, 10000, 12000

```{r}
expression <- c(1000, 3000, 10000, 12000)            # Define expression values

```

### 3. Create a vector of the gene lengths 100, 3000, 200, 1000

```{r}
lengths <- c(100, 3000, 200, 1000)                   # Define gene lengths
```


### 4. Assign the gene names as header for the expression vector

```{r}
names(expression) <- genes   # Assign gene names as names of expression vector
print(expression)            # Display named expression vector

```

### 5. Assign the gene names as header for the gene lengths vector

```{r}
names(lengths) <- genes      # Assign gene names as names of lengths vector
print(lengths)               # Display named lengths vector
```


### 6. Find the longest gene

```{r}
longest_gene <- genes[which.max(lengths)]  # Identify gene with maximum length
print(longest_gene)                        # Display longest gene
```


### 7. Identify genes which have length > 100 and expression > 10000

```{r}
genes_selected <- genes[(lengths > 100) & (expression > 10000)]  # Filter by conditions
print(genes_selected)                                            # Display selected genes
```


---

## Additonal Questions – Derived Calculations

### 1. Create a vector for gene copy number

```{r}
copy_number <- c(10, 1, 50, 12)   # Define gene copy numbers
names(copy_number) <- genes       # Assign gene names as headers
```

### 2. Calculate expression over gene length (length-normalized expression)

```{r}

length_norm_expr <- expression / lengths  # Compute normalized expression
print(length_norm_expr)                   # Display normalized expression

```

### 3. Identify genes with length-normalized expression greater than average

```{r}
genes_high_norm_expr <- genes[length_norm_expr > mean(length_norm_expr)]  # Filter genes above average
print(genes_high_norm_expr)                                               # Display results

```
